Optimizing Abstract Abstract Machines

### 2012-11-14 paper/paper.tex
Added macros for names of types, tried to get some overfull hboxes less full.

Section~\ref{sec:opt} gives a series of simple abstractions and
implementation techniques that, in total, speed up the analysis by
nearly a factor of 500, dropping the analys time to a fraction of a
second.  Figure~\ref{fig:churchtime} shows the step-wise improvement
of the analysis time for this example.

@@

\begin{figure}[t]
\hrule
\vspace{2mm}
\begin{alltt}
(define (id x) x)
(define (f n)
  (if (<= n 1) 1 (* n (f (- n 1)))))
(define (g n)
  (if (<= n 1) 1 (+ (* n n) (g (- n 1)))))
(print (+ ((id f) 3) ((id g) 4)))
\end{alltt}
\hrule
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=4in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.7in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=2.2in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes intepretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

The techniques we propose for optimizing analysis fall into the
following categories:
\begin{enumerate}
\item generate fewer states by avoiding the eager exploration of
  non-deterministic choices that will later collapse into a single
  join point.  We accompish this by applying lazy evaluation
  techniques so that non-determinism is evaluated \emph{by need}.

@@

\item generate states faster.  We accomplish this by better algorithm
  design in the fixpoint computation we use to generate state graphs.
\end{enumerate}
Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for a
small example due to Earl, et
al.~\cite{dvanhorn:Earl2012Introspective}.
By generating significantly fewer states at a significantly faster
rate, we are able to achieve large performance improvements in terms
of both time and space.

@@

\begin{align*}
\ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev{\mexpi{0},\menv,\msto\sqcup[\mlab \mapsto \{\mkont\}],\kar{\mexpi{1},\mlab}}
\\
\co{\kar{\mexp,\menv,\mlab},\mval,\msto} & \machstep
\ev{\mexp,\menv,\msto\sqcup[\mlab^f \mapsto \{\mval\}],\kfn{\mlab^f,\mlab}}
\\
\co{\kfn{\mlab^f,\mlab},\mval,\msto} & \machstep
\ap[_\mlab]{\mvalx{u},\mlab^a,\mkont,\msto\sqcup[\mlab^a \mapsto \{\mval\}]}
\\
\text{ where } &\mkont \in \msto(\maddr), \mvalx{u} \in \msto(\maddr^f)
\\
\ap[_\mlab]{\clos{\mvar,\mexp,\menv},\mlab^a, \msto,\mkont} & \machstep
\ev{\mexp,\menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar \mapsto \msto(\mlab^a)],\mkont}
\\
\ap[_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\\
\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a)\}
\end{align*}

@@

\begin{align*}
force &: Store \times Addr \to \Set(Value) \\
force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
force(\msto,\mval) &= \{\mval\}\\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,\msto\sqcup[\maddr^f \mapsto force(\msto,\mval)],\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], \msto\sqcup[\maddr \mapsto force(\msto, \mval)], \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

@@

\begin{figure}
\begin{align*}
\compile{\_} &: Expr \to Env \times Store \times Kont \to State \\
\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\saddr{\menv(\mvar)},\msto}
\\
\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\mlit,\msto}
\\
\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
\\
&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\\
&\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\end{align*}
\caption{Abstract compilation}
\label{fig:compile}
\end{figure}

@@

We represent changes as $\msdiff \in Store\msdiff = List[Addr \times
  \Set(Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
instead becomes a log addition
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
empty ($\mtlst$) for each step. Applying the changes to the full store
is straightforward:

\begin{align*}
replay &: Store\msdiff \times Store \to Store \\
replay(\mtlst, \msto) &= \msto \\
replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
\end{align*}

@@

Here $concat : \Set(Store\msdiff) \to Store\msdiff$ flattens the lists
of changes to one; order it appends does not matter:
\begin{align*}
concat(\varnothing) &= \mtlst \\
concat(\{\msdiff\}\cup D) &= append(\msdiff, concat(D))
\end{align*}

@@

\begin{alltt}
Value ::= #t | #f | (cons Value Value) | '() | ...
\end{alltt}

>>>>

Section~\ref{sec:opt} gives a series of simple abstractions and
implementation techniques that, in total, speed up the analysis by
nearly a factor of 500, dropping the analysis time to a fraction of a
second.  Figure~\ref{fig:churchtime} shows the step-wise improvement
of the analysis time for this example.

@@

\begin{figure}[t]
\hrule
\vspace{2mm}
\begin{alltt}
(define (id x) x)
(define (f n)
  (if (<= n 1) 1 (* n (f (- n 1)))))
(define (g n)
  (if (<= n 1) 1 (+ (* n n) (g (- n 1)))))
(print (+ ((id f) 3) ((id g) 4)))
\end{alltt}
\hrule
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes intepretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

The techniques we propose for optimizing analysis fall into the
following categories:
\begin{enumerate}
\item generate fewer states by avoiding the eager exploration of
  non-deterministic choices that will later collapse into a single
  join point.  We accomplish this by applying lazy evaluation
  techniques so that non-determinism is evaluated \emph{by need}.

@@

\item generate states faster.  We accomplish this by better algorithm
  design in the fixed-point computation we use to generate state graphs.
\end{enumerate}
Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for a
small example due to Earl, et
al.~\cite{dvanhorn:Earl2012Introspective}.
By generating significantly fewer states at a significantly faster
rate, we are able to achieve large performance improvements in terms
of both time and space.

@@

\newcommand{\ext}[3]{#1\sqcup[#2\mapsto#3]}
%\newcommand{\ext}[3]{ext(#1,#2,#3)}

\begin{align*}
%ext(\msto, \maddr, s) &= \msto \sqcup [\maddr \mapsto s] \\
\ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev{\mexpi{0},\menv,\ext{\msto}{\mlab}{\{\mkont\}},\kar{\mexpi{1},\mlab}}
\\
\co{\kar{\mexp,\menv,\mlab},\mval,\msto} & \machstep
\ev{\mexp,\menv,\ext{\msto}{\mlab^f}{\{\mval\}},\kfn{\mlab^f,\mlab}}
\\
\co{\kfn{\mlab^f,\mlab},\mval,\msto} & \machstep
\ap[_\mlab]{\mvalx{u},\mlab^a,\mkont,\ext{\msto}{\mlab^a}{\{\mval\}}}
\\
\text{ where } &\mkont \in \msto(\maddr), \mvalx{u} \in \msto(\maddr^f)
\\
\ap[_\mlab]{\clos{\mvar,\mexp,\menv},\mlab^a, \msto,\mkont} & \machstep
\ev{\mexp,\menv[\mvar\mapsto\mvar],\ext{\msto}{\mvar}{\msto(\mlab^a)},\mkont}
\\
\ap[_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\\
\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a)\}
\end{align*}

@@

\begin{align*}
force &: \Store \times \Addr \to \Set(\Value) \\
force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
force(\msto,\mval) &= \{\mval\} \\
ext(\msto, \maddr, \mval) &= \msto \sqcup[\maddr \mapsto force(\msto, \mval)] \\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,ext(\msto, \maddr^f,\mval),\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], ext(\msto,\maddr, \mval), \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

@@

\begin{figure}
\begin{align*}
\compile{\_} &: \Expr \to \Env \times \Store \times \Kont \to \State \\
\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\saddr{\menv(\mvar)},\msto}
\\
\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\mlit,\msto}
\\
\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
\\
&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\\
&\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\end{align*}
\caption{Abstract compilation}
\label{fig:compile}
\end{figure}

@@

We represent changes as $\msdiff \in \StoreDelta = List[\Addr \times
  \Set(\Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
instead becomes a log addition
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
empty ($\mtlst$) for each step. Applying the changes to the full store
is straightforward:

\begin{align*}
replay &: \StoreDelta \times \Store \to \Store \\
replay(\mtlst, \msto) &= \msto \\
replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
\end{align*}

@@

Here $concat : \Set(\StoreDelta) \to \StoreDelta$ flattens the lists
of changes to one; order it appends does not matter:
\begin{align*}
concat(\varnothing) &= \mtlst \\
concat(\{\msdiff\}\cup D) &= append(\msdiff, concat(D))
\end{align*}

@@

\newcommand{\alt}{\;\mid\;}
\begin{align*}
\Value &::= \mbox{{\tt \#t}} \alt \mbox{{\tt \#f}} \alt \mbox{{\tt(cons $\Value$ $\Value$)}} \alt \mbox{{\tt '()}} \alt ...
\end{align*}
